Frozen Civil Conflicts

### 2024-03-04 paper.qmd
Draft edits

---
title: "Frozen Civil Conflicts"
subtitle: "The Role of External State Support"
authors:
- name: J.L.A. Krusell
  affiliations:
      - "University of Gothenburg"
date: today
date-format: long
linestretch: 1.25
bibliography: library.bib
filters:
    - lib/luafilter/meta.lua
format:
    pdf:
        template-partials:
            - templates/before-body.tex
            - templates/title.tex
execute:
    cache: true
    echo: false
    warning: false
abstract: |
    Frozen conflicts are among some of the most pernicious, difficult to
    resolve conflicts. Surprisingly, little scholarly attention has been
    paid to the factors influencing their emergence. This paper addresses
    one possible dimension, namely the effect of external state support to
    conflict actors on the probability of an ongoing civil conflict
    becoming frozen. I use an original dataset coding frozen conflicts as
    an outcome among UCDP civil conflict episodes from 1975 to 2019 in
    order to construct a series of bayesian, hierarchical probit models
    motivated by an application of bargaining theory to the domain of
    frozen conflicts. Results indicate that directed support to rebels in
    the absence of similar support to the government side increases the
    probability that a civil conflict episode terminates as a frozen
    conflict. All other treatment conditions --- specifically, directed
    support provided only to the government and directed supported given
    to both sides --- are found to be inconsequential. Taken together,
    these findings help provide a more nuanced understanding of conflict
    termination beyond the common assumption that conflicts either
    conclude through a negotiated settlement or a one-sided military
    victory. Furthermore, this paper also helps shed light on the
    conditions under which actors may prefer a sub-optimal solution,
    allowing a conflict to transition into a state of unstable peace with
    an ever present threat of violent re-escalation, rather than a
    conclusive termination.
---

@@

Accompanying this definition, @smetana_between_2019 lay forth four criteria determining whether a post-conflict period qualifies as a frozen conflict. First, the conflict must be international since it is assumed that frozen conflict are a subset of international conflict featuring state or state-like actors. Second, there needs to be an initial period of active conflict followed by a cessation of large-scale violent hostilities. Isolated, sporadic incidences of violence may continue to occur; however, conflict actors are either unwilling or unable to launch major military offensives. Although not necessary, this may be characterized by the establishment of a general ceasefire and/or presence of external peacekeeping troops. Third, the core issue underlying the conflict needs to remain salient post-conflict. This entails actors maintaining their claims within a dispute. And, finally, there must be an ongoing threat of violent re-escalation. In other words, post-conflict relations should be characterized by mutual hostility and an unwillingness to abandon violence as a political strategy.

These are precise criteria that expand the coverage of frozen conflicts both temporally and geographically, beyond the traditional scope of the post-Soviet separatist states that emerged in the early 1990s following the collapse of the Soviet Union. Furthermore, they hint at the dynamic nature of frozen conflicts --- far from a static condition, when a conflict freezes it may still be characterized by low levels of violence and periodic attempts at negotiations and international mediation. However, I deviate from these criteria in one crucial regard. I relax the assumption that frozen conflicts are a subset of international conflicts and assert that frozen conflicts are a subset of conflict more broadly speaking. In regards to civil conflicts, the @smetana_between_2019 criteria restrict frozen conflicts to *de facto* independent states. This precludes the possibility that non-secessionist conflicts can exhibit the same post-conflict characteristics of unstable peace that define frozen conflicts.

Consider the case of the Second Libyan Civil War. In the aftermath of the overthrow of Muammar Gaddafi in 2011, central political authority fragmented and two competing governments emerged: the internationally recognized and UN sanctioned Government of National Accord (GNA) based in Tripoli and the so-called Tobruk Government comprised of the Libyan House of Representatives elected in the 2014 parliamentary elections aligned with the Libyan National Army under Khalifa Haftar. A ceasefire was signed in 2015 leaving the GNA nominally in control of Tripoli with the Tobruk government in control of the eastern portion of Libya. However, the core contested issue, legitimized control of the central government, remained unresolved with

dyadic frozen conflict

[^libya_ceasefire]: https://www.reuters.com/article/us-libya-security-idUSKBN0KP0VL20150116/

>>>>

---
title: "Frozen Civil Conflicts"
subtitle: "The Role of External State Support"
authors:
- name: J.L.A. Krusell
  email: "joshua.krusell@gu.se"
  affiliations:
      - "University of Gothenburg"
date: today
date-format: long
linestretch: 1.25
bibliography: library.bib
filters:
    - lib/luafilter/meta.lua
format:
    html:
        toc: true
    pdf:
        template-partials:
            - templates/before-body.tex
            - templates/title.tex
execute:
    cache: true
    echo: false
    warning: false
abstract: |
    Frozen conflicts are among some of the most pernicious, difficult to
    resolve conflicts. Surprisingly, little scholarly attention has been
    paid to the factors influencing their emergence. This paper addresses
    one possible dimension, namely the effect of external state support to
    conflict actors on the probability of an ongoing civil conflict
    becoming frozen. I use an original dataset coding frozen conflicts as
    an outcome among UCDP civil conflict episodes from 1975 to 2019 in
    order to construct a series of bayesian, hierarchical probit models
    motivated by an application of bargaining theory to the domain of
    frozen conflicts. Results indicate that directed support to rebels in
    the absence of similar support to the government side increases the
    probability that a civil conflict episode terminates as a frozen
    conflict. All other treatment conditions --- specifically, directed
    support provided only to the government and directed supported given
    to both sides --- are found to be inconsequential. Taken together,
    these findings help provide a more nuanced understanding of conflict
    termination beyond the common assumption that conflicts either
    conclude through a negotiated settlement or a one-sided military
    victory. Furthermore, this paper also helps shed light on the
    conditions under which actors may prefer a sub-optimal solution,
    allowing a conflict to transition into a state of unstable peace with
    an ever present threat of violent re-escalation, rather than a
    conclusive termination.
---

@@

Accompanying this definition, @smetana_between_2019 lay forth four criteria determining whether a post-conflict period qualifies as a frozen conflict. First, the conflict must be international since it is assumed that frozen conflict is a subset of international conflict featuring state or state-like actors. Second, there needs to be an initial period of active, armed conflict followed by a cessation of large-scale violent hostilities. Isolated, sporadic incidences of violence may continue to occur; however, conflict actors are either unwilling or unable to launch major military offensives. Although not necessary, this may be characterized by the establishment of a general ceasefire and/or presence of external peacekeeping troops. Third, the core issue underlying the conflict needs to remain salient post-conflict. This entails actors maintaining their claims within a dispute. And, finally, there must be an ongoing threat of violent re-escalation. In other words, post-conflict relations should be characterized by mutual hostility and an unwillingness to abandon violence as a political strategy.

These are precise criteria that expand the coverage of frozen conflicts both temporally and geographically, beyond the traditional scope of the post-Soviet separatist states that emerged in the early 1990s following the collapse of the Soviet Union. Furthermore, they hint at the dynamic nature of frozen conflicts --- far from a static condition, when a conflict freezes it may still be characterized by low levels of violence and periodic attempts at negotiations and international mediation. However, I deviate from these criteria in one crucial regard. I relax the assumption that frozen conflict is a subset of international conflict and assert that frozen conflict is a subset of conflict more broadly speaking. In regards to civil conflicts, the @smetana_between_2019 criteria restrict frozen conflicts to *de facto* independent states. This precludes the possibility that non-secessionist conflicts can exhibit the same post-conflict characteristics of unstable peace that define frozen conflicts.

Consider the case of the Second Libyan Civil War. In the aftermath of the overthrow of Muammar Gaddafi in 2011, central political authority fragmented and eventually within the midst of a renewed civil war following the contested June 2014 parliamentary elections two competing governments emerged: the internationally recognized Libyan House of Representatives (HoR) elected in 2014 based in Tubruk and in control of much of the Eastern portion of the country with the support of the Libyan National Army under Khalifa Haftar, and the Tripoli-based General National Congress (GNC) originally elected in 2012 and unilaterally self-reinstated in 2014 [@noauthor_libya_nodate-1]. A ceasefire was signed in 2015 followed by the UN brokered Libyan Political Agreement in December that established the Government of National Accord (GNA) with the intention of uniting both factions [@noauthor_libyan_2016]. The agreement was short-lived, by August the HoR had voted to reject the GNA, and cooperation rapidly deteriorated [@noauthor_libya_nodate]. Dyadic hostilities, however, remained at a low-level with only intermittent escalations while focus shifted towards the various Islamist insurgent groups also active in the conflict. Meanwhile, the core contested issue, legitimized control of the central government, remained unresolved with a strong potential for a renewal in high-intensity fighting, which occurred in 2019 with Khalifa Haftar's offensive against Tripoli and the UN-backed GNA [@hibrawi_can_2019].

I contend that the intervening years between the initial ceasefire agreement and the 2019 offensive are best characterized as a dyadic frozen conflict between non-separatist entities within a larger civil war. Although significantly shorter than the typical examples of frozen conflicts involving *de facto* independent separatist states, it satisfies the criteria of an initial period of violence followed by a cessation of large-scale hostilities, albeit with intermittent fighting, and the core issue clearly remained highly salient whilst central government authority was seriously contested.

Finally, it is worth distinguishing frozen conflict from several related concepts, specifically protracted stalemates and enduring internal rivalries. In the case of the former, stalemates during a conflict can potentially endure for extended periods of time and offer conditions similar to those of a frozen conflict. However, they do not necessitate a cessation of active fighting. Instead, they may develop into wars of attrition where actors are unable to achieve an outright military advantage, but continue fighting in the hope of outlasting and "grinding down" the enemy, *i.e.* being better able to endure the prolonged costs of conflict. In the case of the latter, enduring internal rivalries are defined by @derouen_enduring_2008 as "...internal conflicts between a government and an insurgency with at least 10 years of [non-consecutive] armed conflict in which there are at least 25 deaths", an extension of the interstate concept of enduring rivalries [@goertz_enduring_1993] characterized by repeated onsets of militarized disputes. Compared to frozen conflicts, enduring internal rivalries include an explicit temporal requirement and, similar to stalemates, are not restricted to conflicts where fighting has largely stopped. The concept of frozen concepts uniquely captures the post-war condition of unstable peace that characterizes certain conflicts where there is no conclusive termination to the initial episode of fighting.
